\documentclass[11pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\title{Sharp Summary: Bach \& Saremi (2025) -- Binary Denoising and Discrete Langevin}
\author{}
\date{}
\begin{document}
\maketitle

\section*{Problem and scope}
Target distribution $p$ is on the hypercube $\{-1,1\}^d$. The paper asks how to port score/denoising logic to a purely binary space (no Gaussian additive noise).

\section*{Noise model and key identity}
Define Bernoulli sign-flip corruption:
\[
y = x\circ \varepsilon,
\qquad
\mathbb P(\varepsilon_i=1)=\sigma(2\alpha),
\]
with noise level $\alpha\ge 0$.
The noisy law $q_\alpha$ has exponential-tilt form (in $x^\top y$), and the central Tweedie-type identity is
\[
\mathbb E[x\mid y]=\frac{1}{\alpha}\nabla \log q_\alpha(y).
\]
Hence the Hamming-optimal denoiser is
\[
f^*(y)=\operatorname{sign}(\mathbb E[x\mid y]).
\]

\section*{Learning and sampling}
Learning can be done by logistic denoising regression on pairs $(x, x\circ\varepsilon)$, targeting $\mathbb E[x\mid y]$.
Sampling is done on noisy space via discrete Langevin-type Markov kernels using score information, then optional denoising/jump.

The paper proves Wasserstein contractivity/exponential convergence for one-stage and two-stage discrete Langevin variants, and bounds stationary bias relative to target $q_\alpha$; practically, high-noise regimes (small $\alpha$) are easier to mix.

\section*{What this is \emph{not}}
This is not a DDPM forward-time schedule paper and not a Bayes reverse-kernel derivation for finite categorical chains. It is a score-driven MCMC paper on binary spaces.

\section*{Insight for discrete diffusion projects}
Even outside DDPM formalism, it provides a precise message for step-3 style learning:
\begin{itemize}
\item denoising targets can define a valid score surrogate in discrete spaces,
\item noise level controls a geometry-vs-recoverability trade-off,
\item sampling quality depends on both denoiser accuracy and Markov kernel mixing.
\end{itemize}

\section*{Relevance to this repo}
\textbf{Indirect-to-medium.}
Not directly about $Q^t$ Bayes kernels, but very relevant for understanding discrete denoising objectives and Langevin-style alternatives to ancestral reverse-chain sampling.

\end{document}
