\section{Continuous-Time Finite-Chain Formulation and Operator Matching}
\subsection{Continuous-time extension of the finite-state tutorial}
Let $\mathcal X=\{1,\dots,n\}$ and let $(A_t)_{t\in[0,T]}$ be a time-dependent forward generator:
\[
(A_t)_{ij}\ge 0\ (j\neq i),\qquad
(A_t)_{ii}=-\sum_{j\neq i}(A_t)_{ij}.
\]
The forward marginal evolves by the Kolmogorov equation
\[
\frac{d}{dt}h^t=A_t^\top h^t,\qquad h^0\in\Delta_n.
\]
For small step $\tau$, the discrete kernel is
\[
P^t=I+\tau A_t+o(\tau),
\]
which recovers the tutorial recursion $h^{t+\tau}=(P^t)^\top h^t$.

\subsection{Backward generator by infinitesimal Bayes inversion}
Define reverse rates for $i\neq j$ by
\[
(\widetilde A_t)_{ij}
=\frac{h_j^t}{h_i^t}(A_t)_{ji},
\qquad
(\widetilde A_t)_{ii}=-\sum_{j\neq i}(\widetilde A_t)_{ij}.
\]
Then $Q^t=I+\tau \widetilde A_t+o(\tau)$ is the infinitesimal analogue of Bayes reverse kernels.
Moreover, $(h^t)$ is invariant for reverse-time dynamics in the sense
\[
\frac{d}{dt}h^t=-\widetilde A_t^\top h^t.
\]
This is the continuous-time counterpart of $Q^t\in\Kset(h^{t+\tau},h^t)$.

\subsection{Operator matching viewpoint in finite Markov chains}
For test functions $\varphi:\mathcal X\to\mathbb R$, define forward/backward operators
\[
(\mathcal L_t\varphi)(i)=\sum_j (A_t)_{ij}\big(\varphi(j)-\varphi(i)\big),
\]
\[
(\widetilde{\mathcal L}_t\varphi)(i)=\sum_j (\widetilde A_t)_{ij}\big(\varphi(j)-\varphi(i)\big).
\]
The weak-form evolution is
\[
\frac{d}{dt}\langle h^t,\varphi\rangle=\langle h^t,\mathcal L_t\varphi\rangle,
\]
which is the finite-state analogue of the weak continuity-equation formulation used in operator flow matching.

To make operator matching identifiable, one must average over a rich family of test functions.
Let $\Phi\subset \mathbb R^{\mathcal X}$ be a function class and let $\rho$ be a sampling law on $\Phi$.
Given a parametric reverse generator $\widetilde A_\theta(t)$ (equivalently operator $\widetilde{\mathcal L}_\theta$), define
\[
\mathcal J_{\mathrm{op}}(\theta)
\coloneqq
\mathbb E_{t}\,\mathbb E_{\varphi\sim\rho}\,\mathbb E_{i\sim h^t}\Big[
\big((\widetilde{\mathcal L}_\theta)_t\varphi(i)-(\widetilde{\mathcal L}_\star)_t\varphi(i)\big)^2
\Big],
\]
where $\widetilde{\mathcal L}_\star$ is the Bayes reverse operator.
In practice, $\rho$ can be:
\begin{itemize}
\item uniform over a finite basis $\Phi=\{\varphi_m\}_{m=1}^M$ (e.g., canonical basis or low-frequency graph modes), or
\item a random-feature distribution on functions.
\end{itemize}
Then
\[
\mathcal J_{\mathrm{op}}(\theta)\approx \frac1M\sum_{m=1}^M
\mathbb E_t\mathbb E_{i\sim h^t}\!\left[
\big((\widetilde{\mathcal L}_\theta)_t\varphi_m(i)-(\widetilde{\mathcal L}_\star)_t\varphi_m(i)\big)^2
\right].
\]
This is the finite-chain specialization of matching transport operators/vector fields rather than directly matching full conditional distributions.

\paragraph{Link with kernel log-likelihood training.}
For small $\tau$, minimizing transition NLL for kernels
\[
Q_\theta^t=I+\tau \widetilde A_\theta(t)+o(\tau)
\]
and denote off-diagonal reverse rates by
\[
\widetilde a_{\theta,ij}(t)=(\widetilde A_\theta(t))_{ij},\quad i\neq j,
\qquad
\widetilde\lambda_{\theta,i}(t)=\sum_{j\neq i}\widetilde a_{\theta,ij}(t).
\]
For one small step, population NLL (cross-entropy) between $Q_\star^t$ and $Q_\theta^t$ expands as
\[
\mathcal L_t(\theta)
=C_t
+\tau\sum_i h_i^t
\sum_{j\neq i}\Big[
\widetilde a_{\star,ij}\log\frac{\widetilde a_{\star,ij}}{\widetilde a_{\theta,ij}}
-\widetilde a_{\star,ij}
+\widetilde a_{\theta,ij}
\Big]
+o(\tau),
\]
which is the weighted Poisson-KL between reverse rates.
Near the optimum ($\widetilde a_{\theta,ij}\approx \widetilde a_{\star,ij}$),
\[
\mathcal L_t(\theta)-\mathcal L_t(\star)
=\frac{\tau}{2}\sum_i h_i^t\sum_{j\neq i}
\frac{\big(\widetilde a_{\theta,ij}-\widetilde a_{\star,ij}\big)^2}
{\widetilde a_{\star,ij}}
+o\!\left(\tau\|\widetilde A_\theta-\widetilde A_\star\|^2\right),
\]
which is the precise ``quadratic local discrepancy between reverse rates''.

Yes, this can be realized through a suitable choice of test-function distribution $\rho$.
Take $\Phi=\{\mathbf 1_{\{j\}}\}_{j=1}^n$ (canonical indicator functions), uniform $\rho$ on $\Phi$.
Then
\[
(\widetilde{\mathcal L}_t\mathbf 1_{\{j\}})(i)=
\begin{cases}
\widetilde a_{ij}(t), & i\neq j,\\
-\widetilde\lambda_j(t), & i=j,
\end{cases}
\]
so $\mathcal J_{\mathrm{op}}$ becomes a weighted quadratic loss on off-diagonal rates plus escape rates.
Hence, with this $\rho$, operator matching directly controls the same local objects as the small-$\tau$ NLL expansion.
If $\Phi$ is dense/separating, $\mathcal J_{\mathrm{op}}(\theta)=0$ implies $\widetilde A_\theta=\widetilde A_\star$ (on the support of $h^t$), exactly as in the Gaussian continuous-state limit where reverse-kernel NLL reduces to score regression.
